PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-08-11 manuscript_00.Rmd
text updates

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone fovea markers, and propose six new cone fovea and peripheral markers . CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD. This provides the eye community a powerful and quick means to test hypotheses on gene expression across a highly diverse community derived database."

@@

The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al. the adult mouse retina was profiled[@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

@@

Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas[@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries difficult. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].

To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish fovea from peripheral cones in humans.

@@

A custom cutoffs for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200. We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. fovea), sex, and age or developmental stage. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under SubCellType.

@@

The web app at plae.nei.nih.gov can display gene expression in only a few seconds across over a million cells. This speed is only possible due to custom data structures to optimize for data query efficiency. Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories. This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB. For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: gene, cell barcode, and gene count value. This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds. To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields. On user given query, the data is further aggregated to level request. This allows for complicated queries to complete in seconds rather than minutes.

@@

The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data. The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand. However, these resources provide no curation of the metadata and process the data for each study individually. Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical. The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.

@@

Finally we have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse. This plot differs from the Dot Plot view as the expression values are relative instead of absolute. That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells. This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.

@@

Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed. We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#. We find a mean of 
`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`). These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`). The full differential table is made available as supplementary file "differential_expression_table.csv.gz."

@@

@@

```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
neuro_cap <- fig_cap(name='neuro', caption = 'A. Genes identified in Figure 2K of Lu et al. as being differently expressed during the RPC to neurogenic differentiation in human. We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells. The standard error (calculated by DESeq2) is plotted. Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively. B. ZBTB38 .....')
# tic()
# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
#   tbl("diff_testing") %>% 
#   filter(Against  == "RPC",   
#          Group == 'CellType (Predict)', 
#          Base == "Neurogenic Cell") %>% 
#   group_by(Base, Against, Gene) %>% 
#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
#   arrange(-`mean log2FC`) %>%   
#   collect() %>% 
#   filter(!is.na(`mean log2FC`))
# 
# 
# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
# # left_join(gene_tab, by = 'GENEID') %>% 
# # relocate(SYMBOL, GENEID) %>% 
# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
# # arrange(-`mean log2FC`)
#

@@

To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types. We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`). To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types. We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C). We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic. We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C). We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ')` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B).

@@

plot_grid(plot_grid(a,b,c,d + theme(legend.position="none"),e,f,g, nrow = 1),
          region_legend, 
          rel_heights = c(1,0.2), 
          nrow = 2)

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed PRPH2 and RS1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al. proposed CALB1 and GNGT1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

Our differential testing found that while PRPH2, RS1, and GNGT1 had expression differences matching that the findings Yan et al. and Peng et al. published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three. In contrast CALB1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87. We find six more genes with a padj < 0.01: PDE6G, AKAP12, HSPB1, GADD45B, CC2D2A, and RP1L1 (`r fig_cap(name = 'region_fig', display = 'cite')`and `r supTab_cap(name = 'region_table', display = 'cite')`). None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).

@@

@@

## Wealth of data avaibility formats enables custom querying

@@

@@

@@

```{r region_table,  echo = F, message = F, results = 'asis',fig.cap = region_table_cap}
# 5
region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
cat("<div id=\"tbl:region_table\">")
region_table %>% filter(padj < 0.1) %>% 
  arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone macula markers, and propose six new cone macula and peripheral markers . CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD. This provides the eye community a powerful and quick means to test hypotheses on gene expression across a highly diverse community derived database."

@@

The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al. the adult mouse retina was profiled [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

@@

Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries difficult. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].

To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.

@@

A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200. We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under SubCellType.

@@

The web app at plae.nei.nih.gov can display gene expression across over a million cells in only a few seconds. This speed is only possible due to custom data structures that were optimized for data query efficiency. Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories. This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB. For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: gene, cell barcode, and gene count value. This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds. To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields. On user given query, the data is further aggregated to level request. This allows for complicated queries to complete in seconds rather than minutes.

@@

The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app. The codebase for the web app (along with local installation instructins) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data. The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand. However, these resources provide no curation of the metadata and process the data for each study individually. Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical. The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.

@@

We have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse. This plot differs from the Dot Plot view as the expression values are relative instead of absolute. That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells. This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.

@@

Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed. We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#. We find a mean of 
`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`). These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`). The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.

@@

Of course other possibilities for ranking genes are possible. We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization. The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.

@@

```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
neuro_cap <- fig_cap(name='neuro', caption = 'A. Genes proposed by Lu et al. as being significantly differentially expressed during the RPC to neurogenic differentiation in human. We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells. The standard error (calculated by DESeq2) is plotted. Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively. B. Genes from Lu et al. that we replicated with the scEiaD data. C. Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
# tic()
# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
#   tbl("diff_testing") %>% 
#   filter(Against  == "RPC",   
#          Group == 'CellType (Predict)', 
#          Base == "Neurogenic Cell") %>% 
#   group_by(Base, Against, Gene) %>% 
#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
#   arrange(-`mean log2FC`) %>%   
#   collect() %>% 
#   filter(!is.na(`mean log2FC`))
# 
# 
# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
# # left_join(gene_tab, by = 'GENEID') %>% 
# # relocate(SYMBOL, GENEID) %>% 
# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
# # arrange(-`mean log2FC`)
#

@@

To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types. We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`). To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types. We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C). We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic. We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C). We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC. Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.

@@

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al. proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al. and Peng et al. published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three. In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87. We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`). None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).

@@

## Limitations of scEiaD

The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and nearly all cell types. Earlier we proposed well supported cell types. Right now, most of these well supported cell types are in the retina. The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies. Many common human diseases underlie the RPE (e.g. AMD) and the cornea (e.g. Keratoconus) and these tissues will remain a focus for the next update of our database. While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types. For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths. More dramatically, Yan et al. have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a]. While we have found it fairly straightforward to use machine learning to transfer the broad (e.g. rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels. This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes. We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina. Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions. We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.

@@

We have improved on the scEiaD dataset we published in Swamy et al. [@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets. We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data. This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser. We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species. We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.

@@

A great many groups have proposed cell atlases. These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021]. We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas. This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums. Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene cell type signal is stable across an entire community's datasets. We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.

@@

```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}
PLAE_screenshots_supplemental_cap <- supFig_cap(name='PLAE_screenshots_supplemental', caption = 'a. Recreation of Figure 4 directly in the plae.nei.nih.gov web app. b. Identify potential bipolar cell markers with the precomputed differential testing and in c. visualize the top five genes with a Heatmap.')
knitr::include_graphics('fig/plae_sup_fig_screenshots.png')

```{r lu_et_al, echo = F, message = F, fig.cap = lu_et_al_cap}
lu_et_al_cap <- supFig_cap(name='lu_et_al', caption = 'Supplemental figure 2K from Lu et al showing their candidate RPC / Neurogenic differentially expressed genes')
knitr::include_graphics('fig/lu_et_al_fig2k_screenshot.png')

@@

```{r region_table,  echo = F, message = F, results = 'asis',fig.cap = region_table_cap}
# 5
region_table_cap <- supTab_cap(name = 'region_table', caption = 'Macula versus peripheral cone differential gene testing table')
cat("<div id=\"tbl:region_table\">")
region_table %>% filter(padj < 0.1) %>% 
  arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)
